\documentclass[aspectratio=169]{beamer}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{listings}

\definecolor{aoBlue}{RGB}{34,88,153}
\definecolor{aoGreen}{RGB}{40,125,84}
\definecolor{aoGray}{RGB}{246,247,249}
\definecolor{aoInk}{RGB}{38,43,51}

\usetheme{default}
\setbeamertemplate{navigation symbols}{}
\setbeamercolor{title}{fg=aoBlue}
\setbeamercolor{frametitle}{fg=aoBlue}
\setbeamercolor{normal text}{fg=aoInk}
\setbeamercolor{block title}{bg=aoBlue,fg=white}
\setbeamercolor{block body}{bg=aoGray,fg=aoInk}

\lstdefinelanguage{yaml}{
  keywords={version,name,context,providers,command,input_mode,defaults,steps,provider,depends_on,required,inject,mode,position,instruction,input_file,timeout_sec,on,success,goto},
  keywordstyle=\color{aoBlue}\bfseries,
  sensitive=true,
  comment=[l]{\#},
  commentstyle=\color{gray},
  morestring=[b]",
}

\lstset{
  basicstyle=\ttfamily\scriptsize,
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  breaklines=true,
  backgroundcolor=\color{aoGray},
  frame=single,
  rulecolor=\color{aoGray},
}

\title{Workflow Lisp Frontend}
\subtitle{Full drain vs. Ralph loop: implementation quality, throughput, churn, and DSL semantics}
\author{agent-orchestration}
\date{}

\begin{document}

\begin{frame}[plain]
\vspace{0.35cm}
{\Large\color{aoBlue}\textbf{Workflow Lisp Frontend}\par}
\vspace{0.12cm}
{\huge\textbf{Full Drain vs. Ralph Loop}\par}
\vspace{0.2cm}
{\large Implementation quality, throughput, churn, and DSL semantics\par}
\vspace{0.55cm}
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.32\textwidth}
\begin{block}{PL frontend}
\footnotesize
Parsing, ASTs, type checks, macro/module semantics, and lowering into the existing workflow runtime.
\end{block}
\end{column}
\begin{column}{0.32\textwidth}
\begin{block}{Performance}
\footnotesize
Static comparison of full workflow drain coverage against the faster Ralph iteration loop.
\end{block}
\end{column}
\begin{column}{0.32\textwidth}
\begin{block}{Next semantics}
\footnotesize
Adjudicated providers as a step toward workflow-native judging and search loops.
\end{block}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Workflow DSL vocabulary}
\small
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.32\textwidth}
\begin{block}{Workflow}
\footnotesize
A named graph of work. It declares shared context, available providers, and ordered or routed steps.
\end{block}
\end{column}
\begin{column}{0.32\textwidth}
\begin{block}{Step}
\footnotesize
One executable node in the workflow. A step may run a provider, command, assertion, call, loop, branch, or other DSL operation.
\end{block}
\end{column}
\begin{column}{0.32\textwidth}
\begin{block}{Provider}
\footnotesize
A reusable command template for an external worker, such as \texttt{codex exec}. Steps bind to providers and pass them prompts, files, and params.
\end{block}
\end{column}
\end{columns}
\vspace{0.55em}
\begin{block}{In the Ralph loop}
\footnotesize
\texttt{Ralph} is the step. \texttt{codex} is the provider. The workflow routes successful Ralph executions back to Ralph again.
\end{block}
\end{frame}

\begin{frame}[fragile]{Ralph resolves to one Codex command}
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.56\textwidth}
\begin{lstlisting}[language=yaml]
context:
  workflow_model: gpt-5.4
  workflow_effort: high

providers:
  codex:
    command:
      - codex
      - exec
      - --model
      - "${model}"
      - --config
      - "reasoning_effort=${reasoning_effort}"
    input_mode: stdin
    defaults:
      model: "${context.workflow_model}"
      reasoning_effort: "${context.workflow_effort}"
\end{lstlisting}
\end{column}
\begin{column}{0.40\textwidth}
\begin{block}{What this YAML does}
\begin{itemize}
  \item \texttt{provider: codex} selects this command template.
  \item \texttt{context} fills in model and reasoning effort.
  \item The composed prompt is sent to the child process on \texttt{stdin}.
\end{itemize}
\end{block}
\vspace{0.3em}
\begin{block}{Resolved command shape}
\scriptsize
\texttt{codex exec ... --model gpt-5.4 --config reasoning\_effort=high}
\end{block}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile]{The step validates docs and injects paths}
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.52\textwidth}
\begin{lstlisting}[language=yaml,basicstyle=\ttfamily\tiny]
- name: Ralph
  provider: codex
  depends_on:
    required:
      - docs/design/workflow_lisp_frontend_specification.md
      - docs/design/workflow_lisp_frontend_mvp_specification.md
    inject:
      mode: list
      position: prepend
      instruction: "Use these design documents as the contract for this iteration:"
  input_file: workflows/examples/prompts/ralph_lisp_forever/ralph.md
\end{lstlisting}
\end{column}
\begin{column}{0.44\textwidth}
\begin{block}{Runtime dependency contract}
\begin{itemize}
  \item Both docs must exist before the step runs.
  \item Paths resolve relative to the workspace.
  \item \texttt{mode: list} injects filenames only.
  \item The docs are not copied into the prompt.
  \item Codex can read the files on demand from the checkout.
\end{itemize}
\end{block}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile]{Small prompt, unbounded loop}
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.58\textwidth}
\begin{block}{Final stdin prompt}
\begin{lstlisting}[basicstyle=\ttfamily\tiny]
Use these design documents as the contract for this iteration:
  - docs/design/workflow_lisp_frontend_mvp_specification.md
  - docs/design/workflow_lisp_frontend_specification.md

review the lisp mvp and full design docs in context of the progress on its implementation in this repository. Select a non-implemented section of suitable size and draft an implementation plan for it. Then, execute the plan.
\end{lstlisting}
\end{block}
\end{column}
\begin{column}{0.38\textwidth}
\begin{lstlisting}[language=yaml]
  on:
    success:
      goto: Ralph
\end{lstlisting}
\vspace{0.3em}
\begin{block}{Routing}
\small
On success, \texttt{goto: Ralph} loops back to the same step with no iteration cap.
\end{block}
\vspace{0.3em}
\begin{block}{No hidden suffix}
\small
No output contract is appended: Ralph declares no \texttt{expected\_outputs}, \texttt{output\_bundle}, or \texttt{variant\_output}.
\end{block}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile]{Workflow Lisp as a PL frontend}
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.58\textwidth}
\begin{lstlisting}[basicstyle=\ttfamily\tiny,frame=none,backgroundcolor=\color{white}]
Workflow Lisp file
(.orc)
   |
   v
Compiler
   - parses the text
   - checks names and types
   - expands macros/procedures
   - translates Lisp forms
     into workflow operations
   |
   v
Internal workflow representation
   - Core Workflow AST
   - Semantic / executable IR
        ^
        |
YAML workflow (.yaml)
   loads into the same layer
   |
   v
Existing workflow executor
   - runs steps
   - handles loops and resume
   - writes state, ledgers, artifacts
\end{lstlisting}
\end{column}
\begin{column}{0.38\textwidth}
\begin{block}{Why not just a DAG?}
\scriptsize
Workflows are not just one-way task graphs. Steps can branch, retry, pause/resume, repair saved state, and commit outputs. Ralph is the simplest example: on success, it sends itself around again.
\end{block}
\vspace{0.1em}
\begin{block}{What the compiler produces}
\scriptsize
It does not emit machine code or rewrite the workflow as YAML. It builds checked workflow objects that the existing executor runs.
\end{block}
\end{column}
\end{columns}
\end{frame}

\begin{frame}[fragile,shrink=12]{Why a Lisp frontend over YAML?}
\scriptsize
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.49\textwidth}
\begin{block}{YAML: intent plus plumbing}
\begin{lstlisting}[language=yaml,basicstyle=\ttfamily\fontsize{4.25}{4.75}\selectfont]
- name: ExecuteImplementation
  provider: codex
  asset_file: prompts/.../implement.md
  output_bundle:
    path: state/attempt.json

- name: SelectAttempt
  select_variant_output:
    path: state/attempt.json
    discriminant:
      name: implementation_state
      allowed: [COMPLETED, BLOCKED]
    variants:
      COMPLETED:
        fields:
          - name: execution_report_path

- name: PublishReport
  when: implementation_state == COMPLETED
  requires_variant:
    step: SelectAttempt
    value: COMPLETED
  command:
    - python
    - -c
    - |
      target = Path(sys.argv[1])
      pointer = Path(sys.argv[2])
      if not target.exists():
          raise SystemExit(2)
      pointer.write_text(target.as_posix() + "\n")
    - ${steps.SelectAttempt.artifacts.execution_report_path}
    - state/execution_report_path.txt
\end{lstlisting}
\end{block}
\end{column}
\begin{column}{0.49\textwidth}
\begin{block}{Lisp: type and branch together}
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{4.7}{5.15}\selectfont]
;; ReportPath is a checked relpath type.
(defunion ImplementationAttempt
  (COMPLETED
    (execution-report-path ReportPath))
  (BLOCKED
    (progress-report-path ReportPath)
    (blocker-class BlockerClass)))

(let ((attempt
        ;; providers.execute is a build-time alias,
        ;; e.g. the codex provider in this workflow.
        (provider-result providers.execute
          :prompt prompts.implementation.execute
          :returns ImplementationAttempt)))
  (match attempt
    ;; completed is the local name for this case.
    ((COMPLETED completed)
      completed.execution-report-path)
    ((BLOCKED blocked)
      blocked.progress-report-path)))
\end{lstlisting}
\end{block}
\end{column}
\end{columns}

\vspace{0.25em}
\begin{block}{Point}
\fontsize{6.15}{6.85}\selectfont
Same runtime target: provider output, variant selection, and a published report
path. In Lisp, \texttt{defunion} declares the provider result type; each
\texttt{match} arm names the selected case and exposes only that case's fields.
Lowering emits the checked workflow objects that YAML authors currently spell
out by hand.
\end{block}
\end{frame}

\begin{frame}[fragile]{What macros buy}
\scriptsize
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.52\textwidth}
\begin{block}{Macro definition}
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{5.8}{6.4}\selectfont]
(defmacro review-fix-loop
  (name reviewer fixer max)
  `(repeat-until ,name
     :max-cycles ,max
     :body
       ((step Review
          (provider ,reviewer))
        (match Review.decision
          (:approved (done))
          (:needs-fix
            (step Fix
              (provider ,fixer)))))))
\end{lstlisting}
\end{block}
\end{column}
\begin{column}{0.44\textwidth}
\begin{block}{Use site}
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{5.8}{6.4}\selectfont]
(review-fix-loop implementation
  codex-reviewer
  codex-fixer
  3)
\end{lstlisting}
\end{block}
\vspace{0.15em}
\begin{block}{Expands before runtime}
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{5.8}{6.4}\selectfont]
(repeat-until implementation
  :max-cycles 3
  :body
    ((step Review ...)
     (match Review.decision
       (:approved (done))
       (:needs-fix (step Fix ...)))))
\end{lstlisting}
\end{block}
\vspace{0.1em}
\footnotesize
\textbf{Rule:} the executor never runs a macro. Expansion produces normal forms, then diagnostics, typecheck, source maps, and lowering still apply.
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Compiler terms in plain language}
\small
\begin{tabular}{p{0.22\textwidth}p{0.70\textwidth}}
\textbf{Term} & \textbf{Meaning in this workflow frontend} \\
\hline
Syntax forms & Parsed source shapes with enough structure to say what was written and where errors should point. \\
Frontend AST & The Lisp-specific tree after names and forms are understood as workflow concepts. \\
Typed AST & The frontend tree after the compiler checks records, unions, workflow inputs, return values, and allowed operations. \\
Lowering & The translation from Lisp-specific concepts into the shared workflow representation. \\
IR & Intermediate representation: the validated meaning the runtime can execute, resume, and persist. \\
\end{tabular}
\vspace{0.8em}
\begin{block}{Short version}
\footnotesize
The compiler turns human-authored Lisp into a checked workflow meaning. The executor then runs that meaning with the same state, ledger, artifact, and recovery machinery used by YAML workflows.
\end{block}
\end{frame}

\begin{frame}[fragile]{What the AST means}
\small
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.46\textwidth}
\begin{block}{Authored Lisp shape}
\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
(defworkflow RalphDrain
  (step Ralph
    (provider codex)
    (on success Ralph)))
\end{lstlisting}
\end{block}
\end{column}
\begin{column}{0.50\textwidth}
\begin{block}{Simplified tree shape}
\begin{lstlisting}[basicstyle=\ttfamily\scriptsize]
Workflow
  name: RalphDrain
  steps:
    Step
      name: Ralph
      run: Provider(codex)
      on_success: Goto(Ralph)
\end{lstlisting}
\end{block}
\end{column}
\end{columns}
\vspace{0.4em}
\begin{block}{Point}
\footnotesize
The AST is not text to execute directly. It is the workflow split into named parts so the compiler can check it, transform it, and lower it into the runtime's workflow representation.
\end{block}
\end{frame}

\begin{frame}{Full drain vs. Ralph loop: throughput}
\small
\setlength{\tabcolsep}{3pt}
\begin{tabular}{p{0.31\textwidth}p{0.29\textwidth}p{0.29\textwidth}}
\textbf{Metric} & \textbf{Full Lisp drain} & \textbf{Ralph loop} \\
\hline
Elapsed wall time inspected & 31.45 hours & 7.51 hours \\
Completed units & 12 design gaps & 95 Ralph iterations \\
Raw rate & 0.38 design gaps/hour & 12.65 iterations/hour \\
Durable design gaps & 12 recorded gaps & 0 ledger-recorded gaps \\
Functional MVP completion & about 93\% & about 58\% \\
Functional full-design completion & about 72\% & about 27\% \\
Frontend source LOC & 20,675 & 7,130 \\
Direct Lisp test LOC & 10,366 & 4,355 \\
\texttt{.orc} fixtures & 86 & 133 \\
\end{tabular}
\vspace{0.45em}
\begin{block}{Read the units carefully}
The Ralph loop is much faster in raw iteration count, but the units are smaller. The full drain produces larger reviewed design-gap bundles and reaches much deeper into the full Workflow Lisp design.
\end{block}
\vspace{0.1em}
\footnotesize
Source: static inspection of run state, summaries, ledgers, artifacts, source, fixtures, and tests. No tests were run.
\end{frame}

\begin{frame}{Implementation quality comparison}
\small
\begin{tabular}{p{0.34\textwidth}p{0.27\textwidth}p{0.27\textwidth}}
\textbf{Metric} & \textbf{Full Lisp drain} & \textbf{Ralph loop} \\
\hline
Full implementation quality & 7.8 / 10 & 5.8 / 10 \\
Equivalent subset quality & 8.0 / 10 & 6.8 / 10 \\
Spec adherence & 8.1 / 10 & 5.1 / 10 \\
Correctness evidence & 7.6 / 10 & 6.1 / 10 \\
Test coverage & 8.7 / 10 & 6.8 / 10 \\
Maintainability & 6.7 / 10 & 6.5 / 10 \\
Churn severity & 8.5 / 10 & 6.5 / 10 \\
\end{tabular}
\vspace{0.8em}
\begin{block}{Quality readout}
The full drain is heavier and churnier, but it has broader runtime integration, stronger artifact surfaces, deeper test coverage, and more durable completion evidence. The Ralph loop is smaller and fast-moving, but its subset remains less complete and less ledger-backed.
\end{block}
\vspace{0.3em}
\footnotesize
Scores are static-inspection estimates, not test results. Higher quality scores are better; higher churn severity means more repeated repair/rework signal.
\end{frame}

\begin{frame}[shrink=8]{Progress quality: speed vs. churn}
\small
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.48\textwidth}
\begin{block}{Full Lisp drain}
\begin{itemize}
  \item Slower but more complete.
  \item Covers parser, typing, lowering, macros, modules, procedures, stdlib, build artifacts, diagnostics, and migration pieces.
  \item Hottest implementation file appears in 9 execution reports; hottest test appears in 4 reports.
  \item File-repeat evidence is a report-mention proxy; this run has no per-iteration changed-file ledger.
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Ralph loop}
\begin{itemize}
  \item Faster early iteration velocity.
  \item Best suited to small MVP slices and focused regression additions.
  \item Changed-file ledgers show concentrated repeat touches across lowering, expression validation, and related tests.
  \item Hot spots: \texttt{test\_workflow\_lisp\_lowering.py} 55 touches; \texttt{lowering.py} 52 touches.
\end{itemize}
\end{block}
\end{column}
\end{columns}
\vspace{0.6em}
\begin{block}{Repeated-file disparity}
Ralph's hottest source file concentration remains much higher than the full-drain report-mention proxy: 52 changed-file touches versus 9 execution-report mentions. This points to fast local iteration, but also to repeated stabilization in the same subset.
\end{block}
\end{frame}

\begin{frame}[fragile,shrink=6]{Concept: adjudicated provider}
\scriptsize
\begin{columns}[T,totalwidth=\textwidth]
\begin{column}{0.48\textwidth}
\textbf{Current YAML surface}
\begin{lstlisting}[language=yaml,basicstyle=\ttfamily\tiny]
- name: DraftDocument
  adjudicated_provider:
    candidates:
      - id: candidate_a
        provider: demo_candidate_a
      - id: candidate_b
        provider: demo_candidate_b
    evaluator:
      provider: demo_evaluator
      input_file: prompts/evaluate_candidate.md
      evidence_confidentiality: same_trust_boundary
    selection:
      tie_break: candidate_order
    score_ledger_path: artifacts/evaluations/scores.jsonl
  expected_outputs:
    - name: selected_document
      type: relpath
\end{lstlisting}
\end{column}
\begin{column}{0.48\textwidth}
\textbf{Conceptual Lisp surface}
\par{\scriptsize This form would live inside a surrounding \texttt{defworkflow} or \texttt{defproc}; it is not itself a top-level definition.}
\vspace{0.15em}
\begin{lstlisting}[basicstyle=\ttfamily\tiny]
(adjudicated-provider draft-document
  :candidates
    ((candidate-a :provider providers.candidate-a)
     (candidate-b :provider providers.candidate-b))
  :evaluator providers.evaluator
  :evidence same-trust-boundary
  :tie-break candidate-order
  :returns SelectedDocument)
\end{lstlisting}
\end{column}
\end{columns}
\vspace{0.45em}
\begin{block}{Runtime meaning}
\footnotesize
One baseline -> candidate workspaces -> output validation -> evaluator scores -> selected output promotion -> score ledger.
\end{block}
\vspace{0.2em}
\footnotesize
YAML exposes the protocol. A Lisp frontend can name the operation once and lower to the same validated candidate/evaluator/promotion machinery.
\par\medskip
This is an early DSL-semantic step toward LLM-as-a-judge workflows: once judging is a typed operation, later search loops can evolve candidates and use adjudication as the fitness signal.
\end{frame}

\begin{frame}{Toward self-hosted adjudication}
\small
\begin{block}{Current v2.11 semantics}
\begin{itemize}
  \item The evaluator is a provider template plus evaluator prompt/rubric source.
  \item It returns strict score JSON: candidate id, finite score, and summary.
  \item It is not itself a nested workflow, \texttt{call}, \texttt{defproc}, or general DSL expression.
\end{itemize}
\end{block}
\vspace{0.35em}
\begin{block}{Self-hosted direction}
\begin{itemize}
  \item Treat the judge as a typed DSL unit: evidence packet in, score decision out.
  \item Candidate generation, judging, promotion, and score-ledger writing then live in one semantic layer.
  \item Genetic or search-style loops can evolve candidates while adjudication supplies the fitness function.
\end{itemize}
\end{block}
\vspace{0.25em}
\footnotesize
This keeps today's contract honest while showing the path from provider-level adjudication to workflow-native LLM-as-a-judge semantics.
\end{frame}

\end{document}
